Secure Composition of Robust and Optimising Compilers

### 2023-06-17 popl/main.tex
Work on text for baseline def and introduce pstep

For languages used in this paper, a small-step style semantics is used.
That is, there is a step relation between states $\statevar$ and an event $\event$: $\stepto[]{\statevar[_{1}]}{\statevar[_{2}]}{\event}$.
Also, a program's syntax $p$ is encoded in the state $\statevar$ which is written as $p\in\statevar$.
For states $\statevar$, it is assumed that there exists a judgement $\isval{\statevar}$ that determines wether $\statevar$ is a final state or not.
\Cref{fig:starstep:and:nstep}\MK{appendix?} defines the star-step relation $\stepto{\statevar[_{1}]}{\statevar[_{2}]}{\trace}$, essentially modelling a big-step style evaluation, and the $n$-step relation $\stepton{\statevar[_{1}]}{\statevar[_{2}]}{\trace}$ that evaluates a program for $n$ steps.
Both relations filter $\varepsilon$ events.

\begin{figure}[t]
  \judgbox{\stepto{r}{r'}{\trace}}{,,State $\statevar$ steps to final state $\statevar[']$, emitting events $\trace$.''}$\;$\\
  %
  \typerule{star-step-refl}{
    \isval{\statevar}
  }{
    \stepto{\statevar}{\statevar}{\hole{\cdot}}
  }{star-step-refl}
  %
  \typerule{star-step-trans}{
    \event\not=\varepsilon &
    \stepto[]{\statevar}{\statevar[']}{\event} &
    \stepto{\statevar[']}{\statevar['']}{\trace}
  }{
    \stepto{\statevar}{\statevar['']}{\event\cdot\trace}
  }{star-step-trans}
  %
  \typerule{star-step-ign}{
    \stepto[]{\statevar}{\statevar[']}{\varepsilon} &
    \stepto{\statevar[']}{\statevar['']}{\trace}
  }{
    \stepto{\statevar}{\statevar['']}{\trace}
  }{star-step-ign}

  \judgbox{\stepton{\statevar}{\statevar[']}{\trace}}{,,State $r$ does $n$ steps to final state $\statevar[']$, emitting events $\trace$.''}$\;$\\
  %
  \typerule{n-step-refl}{
  }{
    \stepton[0]{\statevar}{\statevar}{\hole{\cdot}}
  }{n-step-refl}
  %
  \typerule{n-step-trans}{
    \event\not=\varepsilon &
    \stepto[]{\statevar}{\statevar[']}{\event} &
    \stepton{\statevar[']}{\statevar['']}{\trace}
  }{
    \stepton[n+1]{\statevar}{\statevar['']}{\event\cdot\trace}
  }{n-step-trans}
  %
  \typerule{n-step-ign}{
    \stepto[]{\statevar}{\statevar[']}{\varepsilon} &
    \stepton{\statevar[']}{\statevar['']}{\trace}
  }{
    \stepton[n+1]{\statevar}{\statevar['']}{\trace}
  }{n-step-ign}

  \caption{Star-step and $n$-step relation for a given step relation $\stepto{\statevar[_{1}]}{\statevar[_{2}]}{\event}$}\label{fig:starstep:and:nstep}
\end{figure}
%
The \emph{behavior} of a whole program $w$ ($\behav{w}$) is a set containing all traces the program may emit during execution.
Note that for deterministic languages, $\behav{w}$ is a singleton set for any $w$.
The behavior of programs gives rise to the definition of satisfaction:

@@

\subsection{Preliminary Definitions}\label{subsec:cs:defs}
\begin{gather*}
  \begin{aligned}
  \mi{(Finals)}~\finalexprnoerr&\bnfdef\ \valueexpr \mid x\hspace{0.5cm}
  \mi{(Results)}~\finalexpr\bnfdef\ \finalexprnoerr \mid stuck\hspace{0.5cm}
  \mi{(Values)}~\valueexpr\bnfdef\ n\\
  \mi{(Expressions)}~\expr&\bnfdef\ \finalexpr \mid \lbinop{\expr[_{1}]}{\expr[_{2}]} \mid \lget{x}{\expr} \mid \lnew{x}{\expr[_{1}]}{\expr[_{2}]} \mid \ldelete{x} \mid \lset{x}{\expr[_{1}]}{\expr[_{2}]} \\
    & \mid \llet{x}{\expr[_{1}]}{\expr[_{2}]} \mid \lreturn{\expr} \mid \lcall{g}{\expr} \mid \lifz{\expr}{\expr[_{1}]}{\expr[_{2}]} \\
    & \text{where}\ \lbinop{}{}\in\left\{+,-,\times,<,/\right\}\\
  \mi{(Functions)}~\function&\bnfdef\ \lfunction{g}{x}{e}\hspace{0.5cm}
  \mi{(Types)}~\type\bnfdef\ \natt\hspace{0.5cm}
  \mi{(Locations)}~\loc\in\nat \\
  \mi{(Eval.~Ctx.)}~\ectx&\bnfdef\ \hole{\cdot} \mid \lbinop{\ectx}{\expr} \mid \lbinop{\valueexpr}{\ectx} \mid \lget{x}{\ectx} \mid \lnew{x}{\ectx}{\expr} \mid \lset{x}{\ectx}{\expr}\\
    & \mid \lset{x}{\valueexpr}{\ectx} \mid \lreturn{\ectx} \mid \lcall{g}{\ectx} \mid \lifz{\ectx}{\expr[_{1}]}{\expr[_{2}]} \\
  \mi{(Poison)}~\poison&\bnfdef\ \poisoned \mid \poisonless \hspace{0.5cm}
  \mi{(Store)}~\store\bnfdef\ \hole{\cdot} \mid \storeel{x}{\loc}{\sandboxtag}{\poison}{n},\store \\
  \mi{(Transfer~Tag)}~\comm&\bnfdef\ \nocomm \mid\ \ctxtocomp \mid\ \comptoctx \hspace{0.5cm}
  \mi{(Heaps)}~\heap\bnfdef\ \hole{\cdot} \mid n,\heap \hspace{0.5cm}
  \mi{(Library)}~\library\bnfdef\ \hole{\cdot} \mid \function,\library \\
  \mi{(Component~Names)}~\commlib&\bnfdef\ \hole{\cdot} \mid g,\commlib \hspace{0.5cm}
  \mi{(Continuation~Stack)}~\ectxstack\bnfdef\ \hole{\cdot} \mid \tup{\ectx}{g},\ectxstack \\
  \mi{(Flow~State)}~\cfstate&\bnfdef\ \tup{\commlib}{\library,\ectxstack} \hspace{0.5cm}
  \mi{(Memory~State)}~\memstate\bnfdef\ \tup{\heap[^{\ctx}]}{\heap[^{\comp}],\store} \hspace{0.5cm}
  \end{aligned}
\end{gather*}

>>>>

For languages used in this paper, a small-step style semantics is used.
That is, there is a step relation between states $\runtimetermvar$ and an event $\event$: $\stepto[]{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\event}$.
Also, a program's syntax $p$ is encoded in the state $\runtimetermvar$ which is written as $p\in\runtimetermvar$.
For states $\runtimetermvar$, it is assumed that there exists a judgement $\isval{\runtimetermvar}$ that determines wether $\runtimetermvar$ is a final state or not.
\Cref{fig:starstep:and:nstep}\MK{appendix?} defines the star-step relation $\stepto{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\trace}$, essentially modelling a big-step style evaluation, and the $n$-step relation $\stepton{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\trace}$ that evaluates a program for $n$ steps.
Both relations filter $\varepsilon$ events.

\begin{figure}[t]
  \judgbox{\stepto{r}{r'}{\trace}}{,,State $\runtimetermvar$ steps to final state $\runtimetermvar[']$, emitting events $\trace$.''}$\;$\\
  %
  \typerule{star-step-refl}{
    \isval{\runtimetermvar}
  }{
    \stepto{\runtimetermvar}{\runtimetermvar}{\hole{\cdot}}
  }{star-step-refl}
  %
  \typerule{star-step-trans}{
    \event\not=\varepsilon &
    \stepto[]{\runtimetermvar}{\runtimetermvar[']}{\event} &
    \stepto{\runtimetermvar[']}{\runtimetermvar['']}{\trace}
  }{
    \stepto{\runtimetermvar}{\runtimetermvar['']}{\event\cdot\trace}
  }{star-step-trans}
  %
  \typerule{star-step-ign}{
    \stepto[]{\runtimetermvar}{\runtimetermvar[']}{\varepsilon} &
    \stepto{\runtimetermvar[']}{\runtimetermvar['']}{\trace}
  }{
    \stepto{\runtimetermvar}{\runtimetermvar['']}{\trace}
  }{star-step-ign}

  \judgbox{\stepton{\runtimetermvar}{\runtimetermvar[']}{\trace}}{,,State $r$ does $n$ steps to final state $\runtimetermvar[']$, emitting events $\trace$.''}$\;$\\
  %
  \typerule{n-step-refl}{
  }{
    \stepton[0]{\runtimetermvar}{\runtimetermvar}{\hole{\cdot}}
  }{n-step-refl}
  %
  \typerule{n-step-trans}{
    \event\not=\varepsilon &
    \stepto[]{\runtimetermvar}{\runtimetermvar[']}{\event} &
    \stepton{\runtimetermvar[']}{\runtimetermvar['']}{\trace}
  }{
    \stepton[n+1]{\runtimetermvar}{\runtimetermvar['']}{\event\cdot\trace}
  }{n-step-trans}
  %
  \typerule{n-step-ign}{
    \stepto[]{\runtimetermvar}{\runtimetermvar[']}{\varepsilon} &
    \stepton{\runtimetermvar[']}{\runtimetermvar['']}{\trace}
  }{
    \stepton[n+1]{\runtimetermvar}{\runtimetermvar['']}{\trace}
  }{n-step-ign}

  \caption{Star-step and $n$-step relation for a given step relation $\stepto{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\event}$}\label{fig:starstep:and:nstep}
\end{figure}
%
The \emph{behavior} of a whole program $w$ ($\behav{w}$) is a set containing all traces the program may emit during execution.
Note that for deterministic languages, $\behav{w}$ is a singleton set for any $w$.
The behavior of programs gives rise to the definition of satisfaction:

@@

\subsection{Preliminary Definitions}\label{subsec:cs:defs}
\begin{gather*}
  \begin{aligned}
  \mi{(Finals)}~\finalexprnoerr&\bnfdef\ \valueexpr \mid x\hspace{0.5cm}
  \mi{(Results)}~\finalexpr\bnfdef\ \finalexprnoerr \mid stuck\hspace{0.5cm}
  \mi{(Values)}~\valueexpr\bnfdef\ n\\
  \mi{(Expressions)}~\expr&\bnfdef\ \finalexpr \mid \lbinop{\expr[_{1}]}{\expr[_{2}]} \mid \lget{x}{\expr} \mid \lnew{x}{\expr[_{1}]}{\expr[_{2}]} \mid \ldelete{x} \mid \lset{x}{\expr[_{1}]}{\expr[_{2}]} \\
    & \mid \llet{x}{\expr[_{1}]}{\expr[_{2}]} \mid \lreturn{\expr} \mid \lcall{g}{\expr} \mid \lifz{\expr}{\expr[_{1}]}{\expr[_{2}]} \\
  \mi{(State)}~\statevar&\bnfdef \tup{\cfstate}{\sandboxtag,\memstate} \hspace{0.5cm}
  \mi{(Runtime~Terms)}~\runtimetermvar\bnfdef\rtt{\statevar}{\expr} \hspace{0.5cm}
    \text{where}\ \lbinop{}{}\in\left\{+,-,\times,<,/\right\}\\
  \mi{(Symbols)}~\function&\bnfdef\ \lfunction{g}{x}{e}\hspace{0.5cm}
  \mi{(Types)}~\type\bnfdef\ \natt\hspace{0.5cm}
  \mi{(Locations)}~\loc\in\nat \\
  \mi{(Eval.~Ctx.)}~\ectx&\bnfdef\ \hole{\cdot} \mid \lbinop{\ectx}{\expr} \mid \lbinop{\valueexpr}{\ectx} \mid \lget{x}{\ectx} \mid \lnew{x}{\ectx}{\expr} \mid \lset{x}{\ectx}{\expr}\\
    & \mid \lset{x}{\valueexpr}{\ectx} \mid \lreturn{\ectx} \mid \lcall{g}{\ectx} \mid \lifz{\ectx}{\expr[_{1}]}{\expr[_{2}]} \\
  \mi{(Poison)}~\poison&\bnfdef\ \poisoned \mid \poisonless \hspace{0.5cm}
  \mi{(Store)}~\store\bnfdef\ \hole{\cdot} \mid \storeel{x}{\loc}{\sandboxtag}{\poison}{n},\store \\
  \mi{(Transfer~Tag)}~\comm&\bnfdef\ \nocomm \mid\ \ctxtocomp \mid\ \comptoctx \hspace{0.5cm}
  \mi{(Heaps)}~\heap\bnfdef\ \hole{\cdot} \mid n,\heap \hspace{0.5cm}
  \mi{(Library)}~\library\bnfdef\ \hole{\cdot} \mid \function,\library \\
  \mi{(Component~Names)}~\commlib&\bnfdef\ \hole{\cdot} \mid g,\commlib \hspace{0.5cm}
  \mi{(Continuation~Stack)}~\ectxstack\bnfdef\ \hole{\cdot} \mid \tup{\ectx}{g},\ectxstack \\
  \mi{(Flow~State)}~\cfstate&\bnfdef\ \tup{\commlib}{\library,\ectxstack} \hspace{0.5cm}
  \mi{(Memory~State)}~\memstate\bnfdef\ \tup{\heap[^{\ctx}]}{\heap[^{\comp}],\store} \\
  \mi{(Programs)}~&\prog{\library_{\ctx}}{\library_{\comp}}\hspace{0.5cm}
  \mi{(Substitutions)}~\substvar\bnfdef \hole{\cdot} \mid \subst{\valueexpr}{x},\gamma \\
  \mi{(Pre\text{--}Events)}~\event[_{b}]&\bnfdef \ev{Alloc}\ \loc\ \valueexpr \mid \ev{Dealloc}\ \loc \mid \ev{Get}\ \loc\ \valueexpr \mid \ev{Set}\ \loc\ \valueexpr\ \valueexpr['] \mid \ev{Call}\ \comm\ g\ \valueexpr \mid \ev{Ret}\ \comm\ \valueexpr\\
    & \mid \ev{Start} \mid \ev{End}\ \valueexpr\hspace{0.5cm}
  \mi{(Events)}~\event\bnfdef \emptyevent \mid \lightning \mid \tup{\event[_{b}]}{\sandboxtag} \\
  \end{aligned}
\end{gather*}

Above is the base syntax of all the programming languages of this paper.
Expressions ($\expr$) must not contain the result $stuck$ unless they are entailed as runtime--term ($\rtt{\statevar}{\expr}$).
States ($\statevar$) have a flow- ($\cfstate$) and a memory state ($\memstate$).
As for the flow state, the list of component--level names ($\commlib$), the list of symbols ($\library$), and a stack of continuations tagged with the name of a symbol ($\ectxstack$) are necessary for calling and returning while emitting the right transfer tag ($\sandboxtag$) depending on whether there is a context switch from context to component ($\comptoctx$), from component to context ($\ctxtocomp$), or whether control flow stays internal to either ($\nocomm$).
The memory state contains both a context--level ($\heap[^{\ctx}]$) and a component--level heap ($\heap[^{comp}]$) as well as a store ($\store$).
Effectively, it is syntactically impossible for a context--level runtime--term to have a pointer into the component--level heap and vice--versa.
Pointers are non-nullable and there is no syntactic way to let them point into the middle of an object.
However, pointers always point to arrays and the individual elements of that array are read- and writeable by means of adding an offset that may be negative.
Heap locations are second--class. %The syntax is constrained as such that one cannot simply guess a heap location.
Stores map identifiers to pointer metadata.
This metadata is a tuple with the location ($\loc$), the control--tag ($\sandboxtag$), poison--tag ($\poison$), and the size of the allocated object ($n$).
The control-tag allows to distinguish whether a location indexes into $\heap[^{\ctx}]$ or $\heap[^{\comp}]$.
The poison--tag marks a location as freed ($\poisoned$) without removing the entry from $\store$.
If one removes an entry, for example, by evaluating $\ldelete{x}$, then there is no way to distinguish between freed pointers and free variables.
The latter should render evaluation as stuck, while the former should emit a $\ev{Dealloc}\ \loc$ event regardless of wether the object a pointer points to is still allocated or not.
Programs ($\prog{\library_{\ctx}}{\library_{\comp}}$) have disjoint sets of symbols, one for context-level ($\library_{\ctx}$) and one for component-level symbols ($\library_{\comp}$).

All programming languages presented in this section have an environmental semantics and all but $\src{L}$ are dynamically typed.
Each evaluation may emit the empty event ($\emptyevent$), a program crash token ($\lightning$), or a pair containing a pre--event ($\event[_{b}]$) and the corresponding control--tag ($\sandboxtag$) indicating whether the event fired during an evaluation of a context or a component.
Note that the reflexive-transitive closure used in this paper behaves like a big-step semantics, where the reflexive case is ony for the empty trace and finals, and the emitted trace does not contain empty events.
Most rules are entirely standard and omitted.
Heaps may need to grow when allocating, which the section writes as $\growh{\heap}{n}$, denoting that $\heap$ grows by $n$ cells.
Indexing into a heap, looking up symbols in libraries, or accessing the metadata of a pointer from a store are all written as a function call.
For example, the value located at the fifth position in heap $\heap$ is written as $\heap(5)$.
The operation to set the fifth position in heap $\heap$ to value $\valueexpr$ is written as $\seth{\heap}{5}{\valueexpr}$ and yields a new heap that is entirely the same to $\heap$ modulo the fifth position.

\begin{center}
  \judgbox{\pstepto{\runtimetermvar}{\runtimetermvar[']}{\event}}{,,Primitive step from runtime--term $\runtimetermvar$ to $\runtimetermvar[']$ emitting $\event$.''}
  \typerule{$e-get-\in$}{
    \sandboxtag=\Omega.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} &
    \loc+n\in\statevar.\heap[^{\sandboxtag}]
  }{
    \pstepto{\rtt{\Omega}{\lget{x}{n}}}{\rtt{\Omega}{\heap[^{\sandboxtag}](\loc+n)}}{\tup{\ev{Get}\ \loc\ n}{\sandboxtag}}
  }{e-get-in}
  %
  \typerule{$e-get-\notin$}{
    \sandboxtag=\Omega.\sandboxtag &
    \statevar.\store(x)=\tup{\loc}{\sandboxtag,\poison,m} &
    \loc+n\notin\statevar.\heap[^{\sandboxtag}]
  }{
    \pstepto{\rtt{\Omega}{\lget{x}{n}}}{\rtt{\Omega}{1729}}{\tup{\ev{Get}\ \loc\ n}{\sandboxtag}}
  }{e-get-notin}
\end{center}

\Cref{tr:e-get-in,tr:e-get-notin} both evaluate a read by looking up the metadata in the store.
But, note that neither performs any bounds-checking and emits a $\ev{Get}\ \loc\ n$ event regardless of whether the access is out of bounds or not.
